Weighted Model Counting Without Parameter Variables

### 2021-03-04 doc/paper3/paper.tex
And more writing

TODO: Mention that formulas, clauses, and models are all treated as sets.

@@

\begin{definition}[Minimum-Cardinality WMC]
  A \emph{minimum-cardinality WMC} instance has same tuple as a WMC instance,
  but its answer is defined to be $\sum_{Y \models \phi\text{, }|Y| = k}
  \prod_{Y \models l} w(l)$, where $k = \min_{Y \models \phi\text{, }Y \ne
    \emptyset} |Y|$, if $k$ exists, otherwise the answer is zero.
\end{definition}

@@

% TODO: incorporate this somewhere
%This additional condition on model
%cardinality becomes necessary because these encodings eliminate clauses of the
%form $p \Rightarrow i$, where $p \in X_P$ is a parameter variable, and $i \in
%X_I$ is an indicator variable. Nonetheless, our transformation algorithm still
%works on such encodings, although the experimental results are discouraging
%because they use approximately twice as many indicator variables. For
%instance, each binary variable of a Bayesian network is encoded using two
%indicator variables while one would suffice.

\section{Pseudo-Boolean Functions}

@@

\section{PBP as a Computational Problem}

\begin{definition}[PBP Instance] \label{def:new_wmc}
  A PBP instance is a tuple $(F, X, \omega)$, where $X$ is the set of variables,
  $F$ is a set of pseudo-Boolean functions $2^X \to \mathbb{R}$, and $\omega \in
  \mathbb{R}$ is the scaling factor.\footnote{We will further restrict $F$ to
    only contain functions that have exactly two values. While this restriction
    is not necessary from a computational point of view, such functions can be
    conveniently represented both in text files and using the syntax proposed in
    this paper.} The \emph{answer} of the instance is $\omega \cdot
  \left(\exists_X\prod_{f \in F}f\right)(\emptyset)$.
\end{definition}

\begin{remark}
  Adding scaling factor $\omega$ to the definition allows us to remove
  clauses that consist entirely of a single parameter variable. The idea of
  extracting some of the structure of the WMC instance into an external
  multiplicative factor originates in one of the Bayesian network encodings
  \cite{DBLP:conf/ecai/BartKLM16}.
\end{remark}

@@

\paragraph{Notes.}
\begin{itemize}
\item The algorithm is based on several observations that will be made precise
  in \cref{thm:correctness}. First, all weights except for $\{w(p) \mid p \in
  X_P\}$ are redundant as they either duplicate an already-defined weight or are
  equal to one. Second, each clause has at most one parameter variable. Third,
  if the parameter variable is negated, we can ignore the clause (this idea
  first appears in the \texttt{cd05} paper \cite{DBLP:conf/ijcai/ChaviraD05}).
\item We formulate our algorithm as a sequel to the WMC encoding procedure
  because the implementations of Bayesian network WMC encodings are all
  closed-source. However, as all transformations in \cref{alg:transformation}
  are local, the algorithm can be efficiently incorporated into a WMC encoding
  algorithm with no noticeable slowdown.
\item No obvious way to do this for \texttt{sbk05} because the roles of
  indicator and parameter (i.e., `chance') variables overlap
  \cite{DBLP:conf/aaai/SangBK05}.
\item Benefits of having this proof in the paper:
  \begin{itemize}
  \item It puts all encodings on a common ground.
  \item It illustrates the convenience of our notation for reasoning about
    (certain types of) pseudo-Boolean functions.
  \item It's too big and too important to be left for the appendix.
  \end{itemize}
\end{itemize}

@@

\begin{algorithm}[t]
  \caption{WMC to PBP transformation}
  \label{alg:transformation}
  \SetKwFunction{rename}{rename}
  \SetKwProg{Fn}{Function}{:}{}
  \KwData{WMC instance $(\phi, X_I, X_P, w)$}
  \KwResult{PBP instance $(F, X, \omega)$}
  $F \gets \emptyset$\;
  $\omega \gets 1$\;
  let $d\colon X_P \to \mathbb{N}$ be defined as $p \mapsto |\{ o \in X_P \mid o
  \le p \}|$\;
  \ForEach{clause $c \in \phi$\label{line:foreach1start}}{
    \uIf{$c \cap X_P = \{ p \}$ for some $p$ \textnormal{\textbf{and}} $w(p) \ne
      1$}{
      \uIf{$|c| = 1$}{$\omega \gets \omega \times w(p)$\;}
      \Else{
        $F \gets F \cup \left\{ \left[ \bigwedge_{l \in c \setminus
              \{p\}} \neg l \right]^{w(p)}_1 \right\}$\;
      }
    }
    \ElseIf{$\{p \mid \neg p \in c\} \cap X_P  = \emptyset$}{
      $F \gets F \cup \{ [c]^1_0 \}$\; \label{line:foreach1end}
    }
  }
  \ForEach{indicator variable $v \in X_I$ \label{line:foreach2start}}{
    \If{$\{[v]_1^p, [\neg v]_1^q\} \subseteq F$ for some $p$ and $q$}{
      $F \gets F \setminus \{ [v]_1^p, [\neg v]_1^q \} \cup \{ [v]_q^p
      \}$\; \label{line:foreach2end}
    }
  }
  replace every variable $v$ in $F$ with $\rename{$v$}$\;
  \Return{$(F, X_I, \omega)$}\;
  \Fn{\rename{$v$}}{
    $S \gets \{u \in X_P \mid u \le v\}$\;
    \lIf{$S = \emptyset$}{\Return{$v$}}
    \Return{$v - d(\max S)$}\;
  }
\end{algorithm}

\subsection{Proof of Correctness}

\paragraph{Notes.}
\begin{itemize}
\item This just says that $p$ is equivalent to a conjunction.
\item The first group of conditions applies to \texttt{d02}, while the second
  group applies to \texttt{bklm16}.
\item For \texttt{cd05} and \texttt{cd06}, condition $2bi$ should be replaced
  with $w(\neg p) = 1$.
\end{itemize}

\paragraph{TODO}
\begin{itemize}
\item check if each condition is actually used. Maybe turn this into a paragraph
  that gives an overview of the proof.
\item condition 1 is necessary in both cases because we're ignoring the weights
  of indicator variables (explicitly acknowledge this)
\item add the \texttt{cd05}/\texttt{cd06} correctness theorem
\end{itemize}

\begin{theorem}[Early Projection
  \cite{DBLP:conf/aaai/DudekPV20,DBLP:conf/cp/DudekPV20},
  verbatim] \label{thm:early}
  Let $X$ and $Y$ be sets of variables. For all functions $f\colon 2^X \to
  \mathbb{R}$ and $g\colon 2^Y \to \mathbb{R}$, if $x \in X \setminus Y$, then
  $\exists_x (f \cdot g) = (\exists_x f) \cdot g$.
\end{theorem}

@@

% TODO: mention that this is similar (but not equal) to the DPMC/ADDMC result
\begin{proposition} \label{prop:equivalence}
  Let $(\phi, X_I, X_P, w)$ be a WMC instance. Then
  \begin{equation}
  \left(\left\{[c]_0^1 \;\middle|\; c \in \phi\right\} \cup \left\{[x]_{w(\neg
        x)}^{w(x)} \;\middle|\; x \in X_I \cup X_P\right\}, X_I \cup X_P,
    1\right) \label{eq:new_wmc}
  \end{equation}
  is a PBP instance with the same answer.
\end{proposition}
\begin{proof}
  Let $f = \prod_{c \in \phi} [c]_0^1$, and $g = \prod_{x \in X_I \cup X_P}
  [x]_{w(\neg x)}^{w(x)}$. Then the answer to the WMC instance
  \eqref{eq:new_wmc} is
  \[
    (\exists_{X_I \cup X_P} fg)(\emptyset) = \sum_{Y \subseteq X_I \cup X_P}
    (fg)(Y) = \sum_{Y \subseteq X_I \cup X_P} f(Y)g(Y)
  \]
  by \cref{lemma:sum}. Note that
  \[
    f(Y) =
    \begin{cases}
      1 & \text{if } Y \models \phi, \\
      0 & \text{otherwise},
    \end{cases}
    \quad
    \text{and}
    \quad
    g(Y) = \prod_{Y \models l} w(l),
  \]
  which means that
  \[
    \sum_{Y \subseteq X_I \cup X_P} f(Y)g(Y) = \sum_{Y \models \phi} \prod_{Y
      \models l} w(l)
  \]
  as required.
\end{proof}

\begin{theorem}[Correctness] \label{thm:correctness}
  \Cref{alg:transformation}, when given a WMC instance $(\phi, X_I, X_P,
  w)$, returns PBP instance with the same answer, provided the following
  conditions are satisfied:
  \begin{enumerate}
  \item for all indicator variables $i \in X_I$, $w(i) = w(\neg i) = 1$,
  \item and either
    \begin{enumerate}
    \item for all parameter variables $p \in X_P$, there is a non-empty family
      of literals $(l_i)_{i=1}^n$ such that \label{condition:d02}
      \begin{enumerate}
      \item $w(\neg p) = 1$,
      \item $l_i \in X_I$ or $\neg l_i \in X_I$ for all $i = 1, \dots, n$,
        \label{condition:equivalence1}
      \item and $\{ c \in \phi \mid p \in c \text{ or } \neg p \in c \} =
        \left\{p \lor \bigvee_{i=1}^n \neg l_i \right\} \cup \{ l_i \lor \neg p
        \mid i = 1, \dots, n \}$; \label{condition:equivalence2}
      \end{enumerate}
    \item or for all parameter variables $p \in X_P$, \label{condition:bklm16}
      \begin{enumerate}
      \item $w(p) + w(\neg p) = 1$,
      \item for any clause $c \in \phi$, $|c \cap X_P| \le 1$,
      \item there is no clause $c \in \phi$ such that $\neg p \in c$,
      \item if $\{ p \} \in \phi$, then there is no clause $c \in \phi$ such
        that $c \ne \{ p \}$ and $p \in c$, \label{cond:just_parameter}
        % 'prime implicants are logically disjoint'
      \item and for any $c, d \in \phi$ such that $c \ne d$, $p \in c$ and $p
        \in d$, $\bigwedge_{l \in c \setminus \{ p \}} \neg l \land \bigwedge_{l
          \in d \setminus \{ p \}} \neg l$ is false. \label{cond:disjoint}
      \end{enumerate}
    \end{enumerate}
  \end{enumerate}
\end{theorem}
\begin{proof}
  By \cref{prop:equivalence},
  \begin{equation}
    \left(\left\{[c]_0^1 \;\middle|\; c \in \phi\right\} \cup \left\{[x]_{w(\neg
          x)}^{w(x)} \;\middle|\; x \in X_I \cup X_P\right\}, X_I \cup X_P,
      1\right) \label{eq:new_wmc2}
  \end{equation}
  is a PBP instance with the same answer as the given WMC instance. By
  \cref{def:new_wmc}, its answer is
  \begin{equation} \label{eq:answer}
    \left(\exists_{X_I \cup X_P} \left( \prod_{c \in \phi} [c]_0^1 \right) \prod_{x \in X_I \cup X_P} [x]_{w(\neg x)}^{w(x)} \right)(\emptyset)
  \end{equation}
  Since both \cref{condition:d02,condition:bklm16} ensure that each clause in
  $\phi$ has at most one parameter variable, we can partition $\phi$ into
  $\phi_* \coloneqq \{c \in \phi \mid \mathtt{Vars}(c) \cap X_P = \emptyset \}$
  and $\phi_p \coloneqq \{ c \in \phi \mid \mathtt{Vars}(c) \cap X_P = \{ p \}
  \}$ for all $p \in X_P$. We can then use \cref{thm:early} to reorder
  \eqref{eq:answer} into
  \[
    \left(\exists_{X_I} \left( \prod_{x \in X_I} [x]_{w(\neg x)}^{w(x)} \right)
      \left( \prod_{c \in \phi_*} [c]_0^1 \right) \prod_{p \in X_P} \exists_p
      [p]_{w(\neg p)}^{w(p)} \prod_{c \in \phi_p} [c]_0^1 \right)(\emptyset).
  \]

  Let us first consider how the unfinished WMC instance $(F, X_I, \omega)$ after
  the loop on \crefrange{line:foreach1start}{line:foreach1end} differs from
  \eqref{eq:new_wmc2}. Note that \cref{alg:transformation} leaves each $c \in
  \phi_*$ unchanged, i.e., adds $[c]_0^1$ to $F$. We can then fix an arbitrary
  $p \in X_P$ and let $F_p$ be the set of functions added to $F$ as a
  replacement of $\phi_p$. It is sufficient to show that
  \begin{equation} \label{eq:to_show}
    \omega \prod_{f \in F_p} f = \exists_p [p]_{w(\neg p)}^{w(p)} \prod_{c \in \phi_p} [c]_0^1.
  \end{equation}
  Note that under \cref{condition:d02},
  \[
    \bigwedge_{c \in \phi_p} c \equiv p \Leftrightarrow \bigwedge_{i=1}^n l_i
  \]
  for some family of indicator variable literals $(l_i)_{i=1}^n$. Thus,
  \[
    \exists_p [p]_{w(\neg p)}^{w(p)} \prod_{c \in \phi_p} [c]_0^1 = \exists_p
    [p]_1^{w(p)} \left[ p \Leftrightarrow \bigwedge_{i=1}^n l_i \right]_0^1.
  \]
  If $w(p) = 1$, then
  \begin{equation} \label{eq:bigsums}
    \exists_p [p]_1^{w(p)} \left[ p \Leftrightarrow \bigwedge_{i=1}^n l_i \right]_0^1 = \left.\left[ p \Leftrightarrow \bigwedge_{i=1}^n l_i \right]_0^1\right|_{p=1} + \left.\left[ p \Leftrightarrow \bigwedge_{i=1}^n l_i \right]_0^1\right|_{p=0}.
  \end{equation}
  Since for any input, $\bigwedge_{i=1}^n l_i$ is either true or false, exactly
  one of the two summands in \cref{eq:bigsums} will be equal to one, and the
  other will be equal to zero, and so
  \[
    \left.\left[ p \Leftrightarrow \bigwedge_{i=1}^n l_i
      \right]_0^1\right|_{p=1} + \left.\left[ p \Leftrightarrow \bigwedge_{i=1}^n
        l_i \right]_0^1\right|_{p=0} = 1,
  \]
  where $1$ is a pseudo-Boolean function that always returns one. On the other
  side of \cref{eq:to_show}, since $F_p = \emptyset$, and $\omega$ is unchanged,
  we get $\omega\prod_{f \in F_p} f = 1$, and so \cref{eq:to_show} is satisfied
  under \cref{condition:d02} when $w(p) = 1$.

  If $w(p) \ne 1$, then
  \[
    F_p = \left\{ \left[ \bigwedge_{i = 1}^n l_i \right]_1^{w(p)} \right\},
  \]
  and $\omega = 1$, and so we want to show that
  \[
    \left[ \bigwedge_{i = 1}^n l_i \right]_1^{w(p)} = \exists_p [p]_1^{w(p)}
    \left[ p \Leftrightarrow \bigwedge_{i=1}^n l_i \right]_0^1,
  \]
  and indeed
  \[
    \exists_p [p]_1^{w(p)} \left[ p \Leftrightarrow \bigwedge_{i=1}^n l_i
    \right]_0^1 = w(p) \cdot \left[ \bigwedge_{i=1}^n l_i \right]_0^1 + \left[
      \bigwedge_{i=1}^n l_i \right]_1^0 = \left[ \bigwedge_{i=1}^n l_i
    \right]_1^{w(p)}.
  \]
  This finishes the proof of the correctness of the first `foreach' loop under
  \cref{condition:d02}.

  Now let us assume \cref{condition:bklm16}. We still want to prove
  \cref{eq:to_show}. If $w(p) = 1$, then $F_p = \emptyset$, and $\omega = 1$,
  and so the left-hand side of \cref{eq:to_show} is equal to one. Then the
  right-hand side is
  \[
    \exists_p [p]_0^1 \prod_{c \in \phi_p} [c]_0^1 = \exists_p \left[ p \land
      \bigwedge_{c \in \phi_p} c \right]_0^1 = \exists_p [p]_0^1 = 0 + 1 = 1
  \]
  since $p \in c$ for every clause $c \in \phi_p$.

>>>>

% TODO: Explain two meanings of entailment in the definition of WMC.

Throughout the paper, we use set theoretic notation for many concepts in logic.
A \emph{clause} is a set of literals that are part of an implicit disjunction.
Similarly, a \emph{formula} in CNF is a set of clauses that are part of an
implicit conjunction. We identify a \emph{model} with a set of variables that
correspond to the positive literals in the model (and all other variables are
the negative literals of the model). We can then define a model's
\emph{cardinality} as the cardinality of this set. For example, let $\phi =
(\neg a \lor b) \land a$ be a propositional formula over variables $a$ and $b$.
Then an equivalent set-theoretic representation of $\phi$ is $\{ \{ \neg a, b
\}, \{ a \} \}$. Any subset of $\{ a, b \}$ is an interpretation of $\phi$,
e.g., $\{ a, b \}$ is a model of $\phi$ (written $\{ a, b \} \models \phi$) of
cardinality two, while $\emptyset$ is an interpretation but not a model.

@@

\begin{definition}
  Let $\phi$ be a propositional formula over a set of variables $X$. Then $Y
  \subseteq X$ is a \emph{minimum-cardinality model} of $\phi$ if $Y \models
  \phi$ and $|Y| \le |Z|$ for all $Z \models \phi$.
\end{definition}

\begin{definition}[Minimum-Cardinality WMC]
  A \emph{minimum-cardinality WMC} instance has same tuple as a WMC instance,
  but its answer is defined to be $\sum_{Y \models \phi\text{, }|Y| = k}
  \prod_{Y \models l} w(l)$ (where $k = \min_{Y \models \phi} |Y|$) if $\phi$ is
  satisfiable, and zero otherwise.
\end{definition}

@@

%This additional condition on model
%cardinality becomes necessary because these encodings eliminate clauses of the
%form $p \Rightarrow i$, where $p \in X_P$ is a parameter variable, and $i \in
%X_I$ is an indicator variable. Nonetheless, our transformation algorithm still
%works on such encodings, although the experimental results are discouraging
%because they use approximately twice as many indicator variables. For
%instance, each binary variable of a Bayesian network is encoded using two
%indicator variables while one would suffice.

\section{Pseudo-Boolean Functions} % TODO: start here
% TODO: need more related literature
% TODO: look at the addmc/background_reading dir

\paragraph{References.} % extensions and applications
\begin{itemize}
\item ADDs \cite{DBLP:journals/fmsd/BaharFGHMPS97}
\item AADDs \cite{DBLP:conf/ijcai/SannerM05}
\item CUDD \cite{somenzi1998cudd}
\item compiling BNs using VE \cite{DBLP:conf/ijcai/ChaviraD07}
\item FOADDs \cite{DBLP:journals/ai/SannerB09}
\item stochastic planning \cite{DBLP:conf/uai/HoeySHB99}
\item XADDs \cite{DBLP:conf/uai/SannerDB11}
\item another connection between BNs and ADDs \cite{DBLP:conf/icml/ZhaoMP15}
\item and one more \cite{DBLP:conf/uai/GogateD11}
\item maybe: \cite{DBLP:conf/aaai/FargierMNS14}
\end{itemize}

@@

\section{Pseudo-Boolean Projection (mostly done)}

We introduce a new type of computational problem called \emph{pseudo-Boolean
  projection} based on two-valued pseudo-Boolean functions. While the same
computational framework can handle any pseudo-Boolean functions, two-valued
functions are particularly convenient because \textsf{DPMC} can be easily
adapted to use them as input, and they are easily representable in both text
files and using the syntax proposed in this paper.

\begin{definition}[PBP Instance] \label{def:new_wmc}
  A PBP instance is a tuple $(F, X, \omega)$, where $X$ is the set of variables,
  $F$ is a set of two-valued pseudo-Boolean functions $2^X \to \mathbb{R}$, and
  $\omega \in \mathbb{R}$ is the scaling factor. The \emph{answer} of the
  instance is $\omega \cdot \left(\exists_X\prod_{f \in F}f\right)(\emptyset)$.
\end{definition}

Adding scaling factor $\omega$ to the definition allows us to remove clauses
that consist entirely of a single parameter variable. The idea of extracting
some of the structure of the WMC instance into an external multiplicative factor
was inspired by the \texttt{bklm16} encoding.

@@

\begin{algorithm}[t]
  \caption{WMC to PBP transformation}
  \label{alg:transformation}
  \SetKwFunction{rename}{rename}
  \SetKwProg{Fn}{Function}{:}{}
  \KwData{WMC (or minimum-cardinality WMC) instance $(\phi, X_I, X_P, w)$}
  \KwResult{PBP instance $(F, X_I, \omega)$}
  $F \gets \emptyset$\;
  $\omega \gets 1$\;
%  let $d\colon X_P \to \mathbb{N}$ be defined as $p \mapsto |\{ o \in X_P \mid o
%  \le p \}|$\;
  \ForEach{clause $c \in \phi$\label{line:foreach1start}}{
    \uIf{$c \cap X_P = \{ p \}$ for some $p$ \textnormal{\textbf{and}} $w(p) \ne
      1$}{
      \uIf{$|c| = 1$}{$\omega \gets \omega \times w(p)$\;}
      \Else{
        $F \gets F \cup \left\{ \left[ \bigwedge_{l \in c \setminus
              \{p\}} \neg l \right]^{w(p)}_1 \right\}$\;
      }
    }
    \ElseIf{$\{p \mid \neg p \in c\} \cap X_P  = \emptyset$}{
      $F \gets F \cup \{ [c]^1_0 \}$\; \label{line:foreach1end}
    }
  }
  \ForEach{indicator variable $v \in X_I$ \label{line:foreach2start}}{
    \If{$\{[v]_1^p, [\neg v]_1^q\} \subseteq F$ for some $p$ and $q$}{
      $F \gets F \setminus \{ [v]_1^p, [\neg v]_1^q \} \cup \{ [v]_q^p
      \}$\; \label{line:foreach2end}
    }
  }
%  replace every variable $v$ in $F$ with $\rename{$v$}$\;
%  \Return{$(F, X_I, \omega)$}\;
%  \Fn{\rename{$v$}}{
%    $S \gets \{u \in X_P \mid u \le v\}$\;
%    \lIf{$S = \emptyset$}{\Return{$v$}}
%    \Return{$v - d(\max S)$}\;
%  }
\end{algorithm}

In this section we describe an algorithm for transforming WMC instances to the
PBP format while removing all parameter variables. The algorithm works on four
out of the five Bayesian network encodings. There is no obvious way to adjust it
to work with \texttt{sbk05} because the roles of indicator and parameter (i.e.,
`chance') variables overlap \cite{DBLP:conf/aaai/SangBK05}. The algorithm is
based on several observations that will be made more precise in
\cref{sec:proof}. First, all weights except for $\{w(p) \mid p \in X_P\}$ are
redundant as they either duplicate an already-defined weight or are equal to
one. Second, each clause has at most one parameter variable. Third, if the
parameter variable is negated, we can ignore the clause (this idea first appears
in the \texttt{cd05} paper \cite{DBLP:conf/ijcai/ChaviraD05}). Note that while
we formulate our algorithm as a sequel to the WMC encoding procedure primarily
because the implementations of Bayesian network WMC encodings are all
closed-source, as all transformations in the algorithm are local, it can be
efficiently incorporated into a WMC encoding algorithm with no slowdown.

The algorithm is listed as \cref{alg:transformation}. The main part of the
algorithms is the first loop that iterates over clauses. If a clause consists of
a single parameter variable, we incorporate it into $\omega$. If a clause is
of the form $\alpha \Rightarrow p$, where $p \in X_P$ and $\alpha$ is a
conjunction of literals over $X_I$, we transform it into a pseudo-Boolean
function $[\alpha]_1^{w(p)}$. If a clause (say, $c \in \phi$) has no parameter
variables, we reformulate it into a pseudo-Boolean function $[c]_0^1$. Finally,
if a clause has negative parameter literals, we skip it.

As all `weighted' pseudo-Boolean functions produced by the first loop are of the
form $[\alpha]_1^p$ (for some $p \in \mathbb{R}$ and formula $\alpha$), the
second loop merges two functions into one whenever $\alpha$ is a literal. Note
that taking into account the order in which clauses are typically generated by
encoding algorithms allows us to do this in linear time (i.e., the two mergeable
functions will be generated one after the other).

@@

\subsection{Correctness Proofs} \label{sec:proof}

In this section we outline key properties that a (WMC or minimum-cardinality
WMC) encoding has to satisfy for \cref{alg:transformation} to output an
equivalent PBP instance. We divide the correctness proof into two theorems:
\cref{thm:correctness} for WMC encodings (i.e., \texttt{bklm16} and
\texttt{d02}) and \cref{thm:mccorrectness} for minimum-cardinality WMC encodings
(i.e., \texttt{cd05} and \texttt{cd06}). We begin by listing some properties of
pseudo-Boolean functions and establishing a canonical transformation from WMC
to PBP.

\begin{theorem}[Early Projection
  \cite{DBLP:conf/aaai/DudekPV20,DBLP:conf/cp/DudekPV20}] \label{thm:early}
  Let $X$ and $Y$ be sets of variables. For all pseudo-Boolean functions
  $f\colon 2^X \to \mathbb{R}$ and $g\colon 2^Y \to \mathbb{R}$, if $x \in X
  \setminus Y$, then $\exists_x (f \cdot g) = (\exists_x f) \cdot g$.
\end{theorem}

@@

\begin{proposition} \label{prop:equivalence}
  Let $(\phi, X_I, X_P, w)$ be a WMC instance. Then
  \begin{equation}
  \left(\left\{[c]_0^1 \;\middle|\; c \in \phi\right\} \cup \left\{[x]_{w(\neg
        x)}^{w(x)} \;\middle|\; x \in X_I \cup X_P\right\}, X_I \cup X_P,
    1\right) \label{eq:new_wmc}
  \end{equation}
  is a PBP instance with the same answer.
\end{proposition}
\begin{proof}
  Let $f = \prod_{c \in \phi} [c]_0^1$, and $g = \prod_{x \in X_I \cup X_P}
  [x]_{w(\neg x)}^{w(x)}$. Then the answer to the WMC instance
  \eqref{eq:new_wmc} is
  \[
    (\exists_{X_I \cup X_P} fg)(\emptyset) = \sum_{Y \subseteq X_I \cup X_P}
    (fg)(Y) = \sum_{Y \subseteq X_I \cup X_P} f(Y)g(Y)
  \]
  by \cref{lemma:sum}. Note that
  \[
    f(Y) =
    \begin{cases}
      1 & \text{if } Y \models \phi, \\
      0 & \text{otherwise},
    \end{cases}
    \quad
    \text{and}
    \quad
    g(Y) = \prod_{Y \models l} w(l),
  \]
  which means that
  \[
    \sum_{Y \subseteq X_I \cup X_P} f(Y)g(Y) = \sum_{Y \models \phi} \prod_{Y
      \models l} w(l)
  \]
  as required.
\end{proof}

\begin{theorem}[Correctness for WMC] \label{thm:correctness}
  \Cref{alg:transformation}, when given a WMC instance $(\phi, X_I, X_P,
  w)$, returns PBP instance with the same answer, provided the following
  conditions are satisfied:
  \begin{enumerate}
  \item for all indicator variables $i \in X_I$, $w(i) = w(\neg i) =
    1$, \label{cond:one}
  \item and either
    \begin{enumerate}
    \item for all parameter variables $p \in X_P$, there is a non-empty family
      of literals $(l_i)_{i=1}^n$ such that \label{cond:d02}
      \begin{enumerate}
      \item $w(\neg p) = 1$,
      \item $l_i \in X_I$ or $\neg l_i \in X_I$ for all $i = 1, \dots, n$,
        \label{condition:equivalence1}
      \item and $\{ c \in \phi \mid p \in c \text{ or } \neg p \in c \} =
        \left\{p \lor \bigvee_{i=1}^n \neg l_i \right\} \cup \{ l_i \lor \neg p
        \mid i = 1, \dots, n \}$; \label{condition:equivalence2}
      \end{enumerate}
    \item or for all parameter variables $p \in X_P$, \label{cond:bklm16}
      \begin{enumerate}
      \item $w(p) + w(\neg p) = 1$,
      \item for any clause $c \in \phi$, $|c \cap X_P| \le 1$, \label{cond:2b2}
      \item there is no clause $c \in \phi$ such that $\neg p \in
        c$, \label{cond:2b3}
      \item if $\{ p \} \in \phi$, then there is no clause $c \in \phi$ such
        that $c \ne \{ p \}$ and $p \in c$, \label{cond:just_parameter}
        % 'prime implicants are logically disjoint'
      \item and for any $c, d \in \phi$ such that $c \ne d$, $p \in c$ and $p
        \in d$, $\bigwedge_{l \in c \setminus \{ p \}} \neg l \land \bigwedge_{l
          \in d \setminus \{ p \}} \neg l$ is false. \label{cond:disjoint}
      \end{enumerate}
    \end{enumerate}
  \end{enumerate}
\end{theorem}

\Cref{cond:one} applies to all four WMC encodings under consideration and
ensures that only parameter variables carry nontrivial weights. \Cref{cond:d02}
(for \texttt{d02}), while spelled out precisely in the theorem, simply states
that each parameter variable is equivalent to a conjunction of indicator
literals. \Cref{cond:bklm16} is for encodings that have implications rather than
equivalences associated with parameter variables (which, in this case, is
\texttt{bklm16}). It ensures that each clause has at most one positive
parameter literal and no negative ones, and that at most one implication clause
per any parameter variable $p \in X_P$ can `force $p$ to be positive'.

\begin{proof}
  By \cref{prop:equivalence},
  \begin{equation}
    \left(\left\{[c]_0^1 \;\middle|\; c \in \phi\right\} \cup \left\{[x]_{w(\neg
          x)}^{w(x)} \;\middle|\; x \in X_I \cup X_P\right\}, X_I \cup X_P,
      1\right) \label{eq:new_wmc2}
  \end{equation}
  is a PBP instance with the same answer as the given WMC instance. By
  \cref{def:new_wmc}, its answer is
  \begin{equation} \label{eq:answer}
    \left(\exists_{X_I \cup X_P} \left( \prod_{c \in \phi} [c]_0^1 \right) \prod_{x \in X_I \cup X_P} [x]_{w(\neg x)}^{w(x)} \right)(\emptyset)
  \end{equation}
  Since both \cref{cond:d02,cond:bklm16} ensure that each clause in
  $\phi$ has at most one parameter variable, we can partition $\phi$ into
  $\phi_* \coloneqq \{c \in \phi \mid \mathtt{Vars}(c) \cap X_P = \emptyset \}$
  and $\phi_p \coloneqq \{ c \in \phi \mid \mathtt{Vars}(c) \cap X_P = \{ p \}
  \}$ for all $p \in X_P$. We can then use \cref{thm:early} to reorder
  \eqref{eq:answer} into
  \[
    \left(\exists_{X_I} \left( \prod_{x \in X_I} [x]_{w(\neg x)}^{w(x)} \right)
      \left( \prod_{c \in \phi_*} [c]_0^1 \right) \prod_{p \in X_P} \exists_p
      [p]_{w(\neg p)}^{w(p)} \prod_{c \in \phi_p} [c]_0^1 \right)(\emptyset).
  \]

  Let us first consider how the unfinished WMC instance $(F, X_I, \omega)$ after
  the loop on \crefrange{line:foreach1start}{line:foreach1end} differs from
  \eqref{eq:new_wmc2}. Note that \cref{alg:transformation} leaves each $c \in
  \phi_*$ unchanged, i.e., adds $[c]_0^1$ to $F$. We can then fix an arbitrary
  $p \in X_P$ and let $F_p$ be the set of functions added to $F$ as a
  replacement of $\phi_p$. It is sufficient to show that
  \begin{equation} \label{eq:to_show}
    \omega \prod_{f \in F_p} f = \exists_p [p]_{w(\neg p)}^{w(p)} \prod_{c \in \phi_p} [c]_0^1.
  \end{equation}
  Note that under \cref{cond:d02},
  \[
    \bigwedge_{c \in \phi_p} c \equiv p \Leftrightarrow \bigwedge_{i=1}^n l_i
  \]
  for some family of indicator variable literals $(l_i)_{i=1}^n$. Thus,
  \[
    \exists_p [p]_{w(\neg p)}^{w(p)} \prod_{c \in \phi_p} [c]_0^1 = \exists_p
    [p]_1^{w(p)} \left[ p \Leftrightarrow \bigwedge_{i=1}^n l_i \right]_0^1.
  \]
  If $w(p) = 1$, then
  \begin{equation} \label{eq:bigsums}
    \exists_p [p]_1^{w(p)} \left[ p \Leftrightarrow \bigwedge_{i=1}^n l_i \right]_0^1 = \left.\left[ p \Leftrightarrow \bigwedge_{i=1}^n l_i \right]_0^1\right|_{p=1} + \left.\left[ p \Leftrightarrow \bigwedge_{i=1}^n l_i \right]_0^1\right|_{p=0}.
  \end{equation}
  Since for any input, $\bigwedge_{i=1}^n l_i$ is either true or false, exactly
  one of the two summands in \cref{eq:bigsums} will be equal to one, and the
  other will be equal to zero, and so
  \[
    \left.\left[ p \Leftrightarrow \bigwedge_{i=1}^n l_i
      \right]_0^1\right|_{p=1} + \left.\left[ p \Leftrightarrow \bigwedge_{i=1}^n
        l_i \right]_0^1\right|_{p=0} = 1,
  \]
  where $1$ is a pseudo-Boolean function that always returns one. On the other
  side of \cref{eq:to_show}, since $F_p = \emptyset$, and $\omega$ is unchanged,
  we get $\omega\prod_{f \in F_p} f = 1$, and so \cref{eq:to_show} is satisfied
  under \cref{cond:d02} when $w(p) = 1$.

  If $w(p) \ne 1$, then
  \[
    F_p = \left\{ \left[ \bigwedge_{i = 1}^n l_i \right]_1^{w(p)} \right\},
  \]
  and $\omega = 1$, and so we want to show that
  \[
    \left[ \bigwedge_{i = 1}^n l_i \right]_1^{w(p)} = \exists_p [p]_1^{w(p)}
    \left[ p \Leftrightarrow \bigwedge_{i=1}^n l_i \right]_0^1,
  \]
  and indeed
  \[
    \exists_p [p]_1^{w(p)} \left[ p \Leftrightarrow \bigwedge_{i=1}^n l_i
    \right]_0^1 = w(p) \cdot \left[ \bigwedge_{i=1}^n l_i \right]_0^1 + \left[
      \bigwedge_{i=1}^n l_i \right]_1^0 = \left[ \bigwedge_{i=1}^n l_i
    \right]_1^{w(p)}.
  \]
  This finishes the proof of the correctness of the first `foreach' loop under
  \cref{cond:d02}.

  Now let us assume \cref{cond:bklm16}. We still want to prove
  \cref{eq:to_show}. If $w(p) = 1$, then $F_p = \emptyset$, and $\omega = 1$,
  and so the left-hand side of \cref{eq:to_show} is equal to one. Then the
  right-hand side is
  \[
    \exists_p [p]_0^1 \prod_{c \in \phi_p} [c]_0^1 = \exists_p \left[ p \land
      \bigwedge_{c \in \phi_p} c \right]_0^1 = \exists_p [p]_0^1 = 0 + 1 = 1
  \]
  since $p \in c$ for every clause $c \in \phi_p$.
